---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# dsb

<!-- badges: start -->
<!-- badges: end -->

This package was developed at [John Tsang's Lab](https://www.niaid.nih.gov/research/john-tsang-phd) by Matt Mulè and Andrew Martins. The package implements our normalization and denoising method for CITEseq data. Technical discussion of how the method works can be found in [the biorxiv preprint](https://biorxiv.org) We utilized the dsb package to normalize CITEseq data reported in this paper [](https://)

In [the biorxiv preprint](https://biorxiv.org), comparing unstained control cells and empty droplets we found the major contribotor to background noise in CITEseq data is unbound antibody captured and sequenced in droplets. DSB corrects for this background by leveraging empty droplets which serve as a "built in" noise measurement in any droplet capture single cell platform (e.g. 10X, dropseq, indrop). 

In addition we define a per-cell denoising covariate to account for the technical component of library size differences between cells which removes spurious cluster formation derived from globally dimcells clustering together. 

## installation 

You can install the released version of dsb in your R session with the command below 

```{r}
# this is analagous to install.packages("package), you need the package devtools to install a package from a github repository like this one. 
require(devtools)
#devtools::install_github(repo = 'MattPM/dsb')

```


## quickstart


```{r example}
# load the example data
library(dsb)
load(cells_citeseq_mtx)
load(empty_drop_citeseq_mtx)

# normalize
normalized_matrix = DSBNormalizeProtein(cell_protein_matrix = cells_citeseq_mtx,
                                        empty_drop_matrix = empty_drop_citeseq_mtx)

```


## The full version (recommended)
By default dsb defines the per cell denoising covariate by fitting a gaussian mixture model to the log + 10 counts of each cell and defining the noise ocvariates the mean. We reccomend including the counts from isotype controls in each cell in the denoising covariates. 

```{r}

# define a vector of the isotype controls in the data 
isotypes = c("Mouse IgG2bkIsotype_PROT", "MouseIgG1kappaisotype_PROT","MouseIgG2akappaisotype_PROT", "RatIgG2bkIsotype_PROT")

normalized_matrix = DSBNormalizeProtein(cell_protein_matrix = cells_citeseq_mtx,
                                        empty_drop_matrix = empty_drop_citeseq_mtx,
                                        use.isotype.control = TRUE,
                                        isotype.control.name.vec = isotypes)


```


# visualize distributions of CD4 and CD8 
plot raw points (overplotted) and points with labeled density distributions (similar to flow)
```{r}
# plot this and avoid plotting by adding a density gradient like a flowjo plot
# this nice density function is from here: https://slowkow.com/notes/ggplot2-color-by-density/
get_density = function(x, y, ...) {
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}

library(ggplot2)
data.plot = normalized_matrix %>% t %>%
  as.data.frame() %>% 
  dplyr::select(CD4_PROT, CD8_PROT) 
data.plot = data.plot %>%   dplyr::mutate(density = get_density(data.plot$CD4_PROT, data.plot$CD8_PROT, n = 100)) 

# plot with and without density gradient
p1 = ggplot(data.plot, aes(x = CD8_PROT, y = CD4_PROT, color = density)) +
  geom_point(size = 0.4) +
  geom_vline(xintercept = 0, color = "red", linetype = 2) + 
  geom_hline(yintercept = 0, color = "red", linetype = 2) + 
  viridis::scale_color_viridis(option = "B") +  
  scale_shape_identity() 
p2 = ggplot(data.plot, aes(x = CD8_PROT, y = CD4_PROT)) +
  geom_point(size = 0.4) +
  geom_vline(xintercept = 0, color = "red", linetype = 2) + 
  geom_hline(yintercept = 0, color = "red", linetype = 2) 

cowplot::plot_grid(p1,p2)

```



The plots above show the actual protein distributions. There is no artificial jitter added to points. 


# How do I get the empty droplets?
  See Vignettes for more info. 
  
  
  


